---
title: "Growth Rates Data Sourcing"
author: "D.Gay"
date: "1/3/2021"
output: html_document
---

```{r message=FALSE, warning=FALSE, results='hide'}
require(OECD)
require(dplyr)
```

### Gathering data

```{r}
dataset_list <- get_datasets()
head(dataset_list, 3)
search_dataset("Quarterly", data = dataset_list)
```

### Proceeding with the 'Quarterly National Accounts' dataset

```{r}
dataset <- "QNA"
dstruc <- get_data_structure(dataset)
str(dstruc, max.level = 1)
dstruc$VAR_DESC
```

```{r}
dstruc$MEASURE
```

We find it in variable "GPSA", which pairs to "Growth rate compared to previous quarter, seasonally adjusted"

### Checking to view the latest period recorded

Looking specifically at the periods directly preceding the pandemic (2018 and 2019 all quarters) and throughout the pandemic (2020 up to 3rd quarter, or current)

We will consider growth rates from 2015-2018 to inform the direction of change for OECD countries leading up to the periods directly preceding the pandemic.

```{r}
tail(dstruc$TIME, 10)
```


### Filtering data

-- 37 OECD nations

-- GPSA: Growth rate compared to previous quarter, seasonally adjusted

-- Quarterly accounts from 2015 through 2020


```{r}
df <- get_dataset("QNA", 
                  filter = c(list(slice_head(dstruc$LOCATION, n=37)$id),
                             "B1_GE", 
                             "GPSA", 
                             "Q"), 
                  start_time = 2015, 
                  end_time = 2020)
```


### Save df to a .csv; all data used for this visualization lives here.

```{r}
df <- within(df, rm(TIME_FORMAT, UNIT, POWERCODE, OBS_STATUS))
names(df)[names(df) == "obsTime"] <- "QUARTER"
names(df)[names(df) == "obsValue"] <- "GROWTH_RATE"
summary(df)
```

